Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-03-19 text/paper2/ms.tex
hogg comments

% commands
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\spitzer}{\project{Spitzer}~}
\newcommand{\given}{\,|\,}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transpose}[1]{{#1}^{\mathsf{T}}}
\newcommand{\inverse}[1]{{#1}^{-1}}

@@

\begin{abstract}
% Context
The dark-matter halo of the Milky Way is expected to be triaxial on large scales
  and filled with substructure on all scales.
It is hoped that streams or shells of stars produced by tidal disruption
  will produce very precise measures of the gravitational potential,
  to test these predictions.
% Aims
We develop a justified method for inferring the gravitational potential
  using precise measurements of stars in tidal streams
  based on an approximate generative model.
Our method does not rely on integrability of the potential;
  we do not know that any good approximation to the large-scale 
  Milky Way potential is integrable.
% Methods: 
Our model is probabilistic, with a likelihood function and priors on the parameters.
It is based on the idea that the stars in a tidal structure
  were close in phase space at some point in the past.
  Model parameters include the present-day phase-space position for each star
  and the progenitor, and thus properly accounts for finite observational uncertainties 
  and missing data.
This method advances in a few key points that have not been jointly addressed in prior work: assumptions of integrability, proper accounting for observational uncertainties, and allowing for missing dimensions.
We test our method with simulated data in a non-integrable triaxial potential, with realistic observational uncertainties for RR Lyrae stars.
Importantly, the fake data were generated with an n-body simulation,
  not with the approximate model underlying our likelihood function.
% Results:
We find that even with a small number (4 to 16) of well-measured stars,
  we can precisely infer properties of the triaxial potential,
  missing phase-space coordinates of the stars,
  and the location of the progenitor object.
The method generalizes to encompass mixtures of independent debris structures,
  or situations in which star membership in structures is unknown.
% [Get more quantitative here?]
% Conclusions: 
\end{abstract}

@@

Early large-scale, cosmological simulations of galaxy formation in the $\Lambda$CDM paradigm suggested that the spherically-averaged density profiles of dark matter halos follow a universal profile across a huge dynamic range in mass \citep{navarro96}. Since then, higher resolution simulations --- both with and without baryons --- have produced dark matter halos that (1) are permeated with substructure on many scales, (2) are triaxial in shape, and (3) have shapes and orientations that vary with radius \citep{dubinski91, jing02, kuhlen07, veraciro11}. Dark-matter-only simulations produce triaxial halos \citep{jing02} with large density fluctuations \citep{zemp09}. Including baryons tends to soften the triaxiality and graininess in the inner galaxy through a combination of dissipative infall \citep{dubinski94} or cooling \citep{bryan13}. These processes combined with the gravity from a baryonic disk or ellipsoid can act to make the inner halo more oblate or spherical, however they do not seem to erase the clumpy, triaxial nature of the outer halo \citep[e.g.,][]{pontzen12}. This can lead to radially-dependent axis ratios, orientation, and smoothness, thus suggesting complicated mass distributions for Milky Way-like galaxies that are not easily represented by integrable potentials.

The bulk of the baryonic matter in galaxies spans roughly 5-10\% of the extent of the host dark matter halo. Hence, the major observable components of a galaxy are sensitive to the inner portion of its host halo's mass distribution. For example, the rotation curves of disk galaxies are excellent probes of the inner mass, especially since matter in disks can be assumed to move on near circular orbits. Conversely, measuring the dark matter distribution at large radii is complicated by the low density of visible tracers, observational difficulties, and assumptions about the smoothness and analytic representation. Around external galaxies, the extended mass distribution has been probed using a variety of approaches \citep[see][for a a complete and detailed review]{courteau13}. For example, the kinematics of tracer populations such as globular clusters or planetary nebulae can be used to derive mass estimates under the assumptions that these satellite systems are relaxed and well-mixed \citep[early investigations include][]{mendez01,cote03}. Simple, parameterized models of both the mass and orbit distribution have been simultaneously constrained using such data \citep[e.g.][]{napolitano11,deason12c}. 
%\citep{romanowsky03,douglas07,romanowsky09,napolitano11,lee11}
Alternatively, the statistical properties of gravitationally lensed background sources around a galaxy can be used to constrain the \emph{projected} shape, orientation, and radial profile of mass \citep[as done by the Lens Structure and Dynamics Survey described in][]{koopmans02}. Of course, lensing reconstructions can only be performed for galaxies which closely intersect our line of sight to background sources, but the advent of large photometric catalogues has allowed automatic searches for such chance alignments and significant increases in the number of objects studied in this way \citep[e.g. the Sloan Lens ACS Survey, see][]{bolton06}.

@@

Tidal streams are dynamically cold systems --- debris typically have small distributions of energy and angular momentum --- and thus require orders of magnitude fewer tracers than a random sample to get constraints of comparable accuracy. For example, in the simplest case we might assume that debris stars are actually still on the same orbit as their progenitor system \citep[a \emph{wrong} assumption, see e.g.,][]{eyre11}. Then we can imagine measuring the full-space velocities ${\bf v}$ at different points ${\bf x}$ in the structure (e.g., along a stream), which would give us a direct measure of differences in a potential, $\Phi$, without running a full model of the disruption. An extension of this idea is to fit a single orbit to observed debris \citep[e.g.,][]{koposov10, deg13}. Though this is known to be incorrect and leads to biases in inferred properties of the underlying potential \citep[e.g.,][]{eyre11, lux13, sanders13a}, \cite{deg14} and \cite{lux13} have used orbit fitting to demonstrate the power of combining multiple streams in dynamical inference. To account for the offset between the orbit of the progenitor and the orbits of the debris stars, methods have been proposed that model the debris as an orbit plus some dispersion or offset, either in phase space \citep[e.g.,][]{eyre09a, varghese11, kuepper12} or action-angle coordinates \citep{eyre11, sanders13b, bovy14, sanders14}. Other statistical methods have been proposed \citep[][]{penarrubia12, sanderson14} that may prove powerful when applied to, e.g., data from the \gaia\, mission, where full 6D coordinates will be known for large samples of stars in the halo --- and therefore many debris structures --- but stream membership is not known for all stars.

We identify a number of problems with approximate models; each method  mentioned above solves only a subset of these issues:
\begin{enumerate}
	\item The known debris structures are $\gtrsim$10 kpc from the Sun, thus while overdensities in the halo are less contaminated by background, the constituent stars often have incomplete or poorly measured phase-space information. Thus it is critical for any approximate model of tidal debris to incorporate observational uncertainties and missing dimensions in a consistent and justified way. 
	\item The potential of the inner galaxy is plausibly well-described by a (set of) nearly-integrable potentials, thus the dynamics of collisionless systems is conceivably captured by a transformation to some integral-of-motion-space, e.g., action-angle coordinates. However there is large uncertainty in the radial profile, shape, orientation, and graininess of the outer halo and the constancy of these parameters over distance. To test the degree to which the outer halo is well-fit by integrable potentials, it is problematic to use methods that \emph{assume} integrability. 
	\item Near-future photometric surveys such as \gaia\, and LSST will likely discover many new streams and kinematic associations of stars. Potential recovery methods should be able to simultaneously model multiple streams and incorporate other dynamical constraints.
	\item Matching generated models to observed stream densities is difficult: models that rely on this must account for observational biases and properly model the underlying background of halo stars. 
	\item Full N-body models are expensive to run. Incorporating an n-body model into a general model and then performing a parameter search would be computationally intensive and, presently, intractable. 
\end{enumerate}

@@

For all experiments below, we use a Markov Chain Monte Carlo (MCMC) algorithm to sample from the posterior probability distribution given by our model. We use an affine-invariant ensemble sampler \citep{goodman10} implemented in the Python programming language \citep{foremanmackey13} and run the walkers in parallel using the message passing interface (MPI). We compute the autocorrelation time for each sampled parameter (using \texttt{ACOR}\footnote{www.math.nyu.edu/faculty/goodman/software/acor/}$^{,}$\footnote{https://github.com/dfm/acor}) and thin the chains by taking every $\mathrm{median}(t_{\rm acor})$ sample to ensure the samples are close to independent.

@@

We first test the model using eight particles (stars) --- four from the leading tail, and four from the trailing tail --- randomly sampled from the $2.5\times10^8~M_\odot$ simulation (Section~\ref{sec:sims}), assuming perfect observational data for both the stars and the progenitor. The stars were required to have been stripped after the first pericentric passage and have a present-day distance within $50$ kpc of the Sun. Figure~\ref{fig:sims} (second column from right) shows the eight randomly chosen stars (black crosses) in Galactic coordinates, over-plotted on all other simulated particles (grey points).

>>>>

% commands
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\spitzer}{\project{Spitzer}~}
\newcommand{\given}{\,|\,}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transpose}[1]{{#1}^{\mathsf{T}}}
\newcommand{\inverse}[1]{{#1}^{-1}}
%\newcommand{\project}[1]{{\sffamily #1}}

@@

\begin{abstract}
% Context
The dark-matter halo of the Milky Way is expected to be triaxial on large scales
  and filled with substructure.
It is hoped that streams or shells of stars produced by tidal disruption
  will provide very precise measures of the gravitational potential,
  to test these predictions.
% Aims
We develop a justified method for inferring the gravitational potential
  using precise measurements of stars in tidal streams
  based on an approximate generative model.
Our method does not rely on integrability of the potential;
  we do not know that any good approximation to the large-scale 
  Milky Way potential is integrable.
% Methods: 
Our model is probabilistic, with a likelihood function and priors on the parameters.
It is based on the idea that the stars in a tidal structure
  were close in phase space at some point in the past.
  Model parameters include the present-day phase-space position for each star
  and the progenitor, and thus properly accounts for finite observational uncertainties 
  and missing data.
This method is new in that it makes no assumption about the form of the Milky Way 
  potential, accounts properly for observational uncertainties, and allows for missing data
  dimensions.
We test our method with simulated data in a non-integrable triaxial potential, with realistic 
  observational uncertainties for RR Lyrae stars.
Importantly, the fake data were generated with an n-body simulation,
  not with the approximate model underlying our likelihood function.
% Results:
We find that even with a small number (8 to 16) of well-measured stream stars,
  we can precisely infer properties of the triaxial potential,
  missing phase-space coordinates of the stars,
  and the location of the progenitor object.
The method generalizes to encompass mixtures of independent debris structures,
  or situations in which star membership in structures is unknown.
% [Get more quantitative here?]
% Conclusions: 
\end{abstract}

@@

Early large-scale, cosmological simulations of galaxy formation in the $\Lambda$CDM paradigm suggested that the spherically-averaged density profiles of dark matter halos follow a universal profile across a large dynamic range in mass \citep{navarro96}. Since then, higher resolution simulations --- both with and without baryons --- have produced dark matter halos that (1) are permeated with substructure on many scales, (2) are triaxial in shape, and (3) have shapes and orientations that vary with radius \citep{dubinski91, jing02, kuhlen07, veraciro11}. Dark-matter-only simulations produce triaxial halos \citep{jing02} with large density fluctuations \citep{zemp09}. Inclusion of baryons tends to soften the triaxiality and graininess in the inner galaxy through a combination of dissipative infall \citep{dubinski94} or cooling \citep{bryan13}. These processes combined with the gravity from a baryonic disk or ellipsoid can act to make the inner halo more oblate or spherical, however they do not seem to erase the clumpy, triaxial nature of the outer halo \citep[e.g.,][]{pontzen12}. This can lead to radially-dependent axis ratios, orientation, and smoothness, thus suggesting complicated mass distributions for Milky Way-like galaxies that are not easily represented by integrable potentials.

The bulk of the baryonic matter in galaxies spans roughly 5-10\% of the extent of the host dark matter halo. Hence, the most observable components of a galaxy are sensitive to the inner portion of the host halos mass distribution. For example, the rotation curves of disk galaxies trace the inner mass, especially since matter in disks can be assumed to move on nearly circular orbits. Conversely, measurements of the dark matter distribution at large radii is complicated by the low density of visible tracers, observational difficulties, and assumptions about the smoothness and analytic form. Around external galaxies, the extended mass distribution has been studied using a variety of approaches \citep[see][for a a complete and detailed review]{courteau13}. For example, the kinematics of tracer populations such as globular clusters or planetary nebulae can be used to derive mass estimates under the assumptions that these satellite systems are relaxed and well-mixed \citep[early investigations include][]{mendez01,cote03}. Simple, parameterized models of both the mass and orbit distribution have been simultaneously constrained using such data \citep[e.g.][]{napolitano11,deason12c}. 
%\citep{romanowsky03,douglas07,romanowsky09,napolitano11,lee11}
Alternatively, the statistical properties of gravitationally lensed background sources around a galaxy can be used to constrain the \emph{projected} shape, orientation, and radial profile of mass \citep[see, for example, the Lens Structure and Dynamics Survey described in][]{koopmans02}. Of course, lensing reconstructions can only be performed for galaxies which closely intersect our line of sight to background sources, but the advent of large photometric catalogues has allowed automatic searches for such chance alignments and significant increases in the number of objects studied in this way \citep[e.g. the Sloan Lens ACS Survey, see][]{bolton06}.

@@

Tidal streams are dynamically cold systems --- debris typically have small distributions of energy and angular momentum --- and thus require orders of magnitude fewer tracers than a random sample to get constraints of comparable accuracy to, e.g., Jeans analysis. For example, in the simplest case we might assume that debris stars are actually still on the same orbit as their progenitor system \citep[a \emph{wrong} assumption, see e.g.,][]{eyre11}. Then we can imagine measuring the full-space velocities ${\bf v}$ at different points ${\bf x}$ in the structure (e.g., along a stream), which would give us a direct measure of differences in a potential, $\Phi$, without running a full model of the disruption. An extension of this idea is to fit a single orbit to observed debris \citep[e.g.,][]{koposov10, deg13}. Though this is known to be incorrect and leads to biases in inferred properties of the underlying potential \citep[e.g.,][]{eyre11, lux13, sanders13a}, \cite{deg14} and \cite{lux13} have used orbit fitting to demonstrate the power of combining multiple streams in dynamical inference. To account for the offset between the orbit of the progenitor and the orbits of the debris stars, methods have been proposed that model the debris as an orbit plus some dispersion or offset, either in phase space \citep[e.g.,][]{eyre09a, varghese11, kuepper12} or action-angle coordinates \citep{eyre11, sanders13b, bovy14, sanders14}. We will \emph{not} assume that streams follow the orbit of the progenitor. Other statistical methods have been proposed \citep[][]{penarrubia12, apw13, sanderson14} that may prove powerful when applied to, e.g., data from the \gaia\, mission, where full 6D coordinates will be known for large samples of stars in the halo --- and therefore many debris structures --- but stream membership is not known for all stars.

We identify a number of problems with existing approximate models; each method mentioned above solves only a subset of these issues:
\begin{enumerate}
	\item The known debris structures are $\gtrsim$10 kpc from the Sun, thus while overdensities in the halo are less contaminated by background, the constituent stars often have incomplete or poorly measured phase-space information. Thus it is critical for any approximate model of tidal debris to incorporate observational uncertainties and missing dimensions in a consistent and justified way. 
	\item The potential of the inner galaxy is conceivably well-described by a (set of) nearly-integrable potentials, thus the dynamics of collisionless systems is conceivably captured by a transformation to some integral-of-motion-space, e.g., action-angle coordinates. However there is large uncertainty in the radial profile, shape, orientation, and graininess of the outer halo and the constancy of these parameters over distance. To test the degree to which the outer halo is well-fit by integrable potentials, it is problematic to use methods that \emph{assume} integrability. 
	\item Near-future photometric surveys such as \gaia\, and the \project{LSST} will likely discover many new streams and kinematic associations of stars. Potential recovery methods should be able to simultaneously model multiple streams and incorporate other dynamical constraints.
	\item Matching generated models to observed stream densities is difficult: models that rely on this must account for observational biases and properly model the underlying background of halo stars. 
	\item Full N-body models are expensive to run. Incorporating an n-body model into a general model and then performing a parameter search would be computationally intensive and, presently, intractable. 
\end{enumerate}

@@

For all experiments below, we use a Markov Chain Monte Carlo (MCMC) algorithm to sample from the posterior probability distribution given by our model. Standard MCMC algorithms (e.g., Metropolis-Hastings) update a single chain while exploring parameter space. We instead use an affine-invariant ``ensemble'' sampler \citep{goodman10} that, each step in parameter space, updates the positions of many ``walkers''. This algorithm is implemented in the \project{Python} programming language \citep{foremanmackey13} and runs naturally in a parallelized environment, e.g., the message passing interface (MPI). We compute the autocorrelation time for each sampled parameter (using \texttt{ACOR}\footnote{www.math.nyu.edu/faculty/goodman/software/acor/}$^{,}$\footnote{https://github.com/dfm/acor}) and thin the chains by taking every $\mathrm{median}(t_{\rm acor})$ sample to ensure the samples are close to independent.

@@

We first test the model using eight particles (stars) --- four from the leading tail, and four from the trailing tail --- randomly sampled from the $2.5\times10^8~M_\odot$ simulation (Section~\ref{sec:sims}), assuming perfect knowledge of the observed 6D positions for both the stars and the progenitor. The stars were required to have been stripped after the first pericentric passage and have a present-day distance within $50$ kpc of the Sun. Figure~\ref{fig:sims} (second column from right) shows the eight randomly chosen stars (black crosses) in Galactic coordinates, over-plotted on all other simulated particles (grey points).
